\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-16 analysis/paper/paper.tex
add abstract and plot captions

\Abstract{
The abstract of the article. *SRB and MT contributed equally to this
work.
}

\Keywords{survival analysis, absolute risk, data visualization}
\Plainkeywords{survival analysis, absolute risk, data visualization}

@@

Using the framework of logistic regression for analysing time-to-event
data opens the door to an extensive array of powerful modeling tools.
Indeed, lasso and elastic-net regression can then be used to select
variables that are associated with the hazard function. Logistic
regression can also be replaced by multinomial regression to analyse
data with multiple competing events.

@@

\caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset. The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model.}\label{fig:plot-stratified-erspc-data}
\end{figure}
\end{CodeChunk}

@@

As above, we can look at the population-time plot for a graphical
summary of the event incidence. Here, we colour the exposed person-time
(i.e.~after transplant) in a darker shade of gray. As we can see, most
events occur early during the follow-up period, and therefore we do not
expect the hazard to be constant.

\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1} \end{center}

@@

\begin{CodeInput}
R> plot(fit3, hazard.params = list(xvar = "futime",
+                                 by = "exposure",
+                                 alpha = 0.05,
+                                 ylab = "Hazard",
+                                 data = cb_data))
\end{CodeInput}

\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1} \end{center}

The non-proportionality seems to be more pronounced at the beginning of
follow-up than the end. Finally, we can turn these estimates of the
hazard function into estimates of the cumulative incidence functions.

@@

\begin{CodeInput}
R> 
R> plot(absrisk,
+      id.names = c("No Tx", "Tx")) + 
+   ylab("Cumulative Incidence") +
+   xlab("Follow-up time (in days)") +
+   paper_gg_theme
\end{CodeInput}

\begin{center}\includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1} \end{center}

@@

This report was generated on 2020-09-15 17:46:19 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.7.0               dotwhisker_0.5.0         
#>  [3] timereg_1.9.7             glue_1.4.2               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.2              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.1.2               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.9          
#> [15] visreg_2.7.0              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.2-3           
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.2.1          glmnet_4.0-2             
#> [25] Matrix_1.2-18             data.table_1.13.0        
#> [27] cowplot_1.1.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_2.0.1     base64enc_0.1-3    
#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       highr_0.8          
#> [55] foreach_1.5.0       checkmate_2.0.0     lava_1.6.7         
#> [58] shape_1.4.4         rlang_0.4.7         pkgconfig_2.0.3    
#> [61] matrixStats_0.56.0  evaluate_0.14       lattice_0.20-41    
#> [64] labeling_0.3        htmlwidgets_1.5.1   cmprsk_2.2-10      
#> [67] tidyselect_1.1.0    deSolve_1.28        plyr_1.8.6         
#> [70] R6_2.4.1            generics_0.0.2      Hmisc_4.4-1        
#> [73] multcomp_1.4-13     DBI_1.1.0           withr_2.2.0        
#> [76] pillar_1.4.6        haven_2.3.1         foreign_0.8-80     
#> [79] mgcv_1.8-33         nnet_7.3-14         mstate_0.2.12      
#> [82] modelr_0.1.8        crayon_1.3.4        rmarkdown_2.3      
#> [85] jpeg_0.1-8.1        grid_4.0.2          readxl_1.3.1       
#> [88] blob_1.2.1          reprex_0.3.0        digest_0.6.25      
#> [91] webshot_0.5.2       numDeriv_2016.8-1.1 stats4_4.0.2       
#> [94] munsell_0.5.0       viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    master /Users/maxturgeon/Documents/git_repos/cbpaper
#> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
#> Head:     [ddc5b48] 2020-09-14: Merge pull request #29 from sahirbhatnagar/editSahir
\end{CodeOutput}
\end{CodeChunk}

>>>>

\Abstract{
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given a
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. But to estimate
risks, the baseline hazard needs to be estimated separately using a
non-parametric method. This apparoch leads to step-wise estimates of the
cumulative incidence that are difficult to interpret. Using case-base
sampling, Hanley \& Miettinen explained how the parametric hazard
functions can be estimated using logistic regression. Their approach
naturally leads to estimates of the cumulative incidence that are
smooth-in-time.\\
In this paper, we present the casebase R package, which implements the
ideas of Hanley \& Miettinen through a simple and flexible interface. We
describe how the case-base framework can also be used in more complex
settings: competing risks, time-varying exposure, and variable
selection. Our package also includes an extensive array of visualization
tools to complement the analysis of time-to-event data. We illustrate
all these features through four different case studies.\\
*SRB and MT contributed equally to this work.
}

\Keywords{survival analysis, absolute risk, \proglang{R}, data
visualization}
\Plainkeywords{survival analysis, absolute risk, R, data visualization}

@@

Using the framework of logistic regression for analyzing time-to-event
data opens the door to an extensive array of powerful modeling tools.
Indeed, lasso and elastic-net regression can then be used to select
variables that are associated with the hazard function. Logistic
regression can also be replaced by multinomial regression to analyse
data with multiple competing events.

@@

\caption[Population time plots for both treatment arms in the ERSPC dataset]{Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model.}\label{fig:plot-stratified-erspc-data}
\end{figure}
\end{CodeChunk}

@@

As above, we can look at the population-time plot for a graphical
summary of the event incidence (see Figure \ref{fig:cs4PopTime}. Here,
we colour the exposed person-time (i.e.~after transplant) in a darker
shade of gray. As we can see, most events occur early during the
follow-up period, and therefore we do not expect the hazard to be
constant.

{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-poptime-1}

\caption{\label{fig:cs4PopTime} Population-time for the Stanford Transplant study. The darker shade of gray corresponds to the exposed person-time.}\label{fig:stanford-poptime}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> plot(fit3, hazard.params = list(xvar = "futime",
+                                 by = "exposure",
+                                 alpha = 0.05,
+                                 ylab = "Hazard",
+                                 data = cb_data))
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-hazard-1}

\caption{\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals.}\label{fig:stanford-hazard}
\end{figure}
\end{CodeChunk}

Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems
to be more pronounced at the beginning of follow-up than the end.
Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the
hazard function into estimates of the cumulative incidence functions.

@@

\begin{CodeInput}
R> 
R> plot(absrisk,
+      id.names = c("No Tx", "Tx")) + 
+   ylab("Cumulative Incidence") +
+   xlab("Follow-up time (in days)") +
+   paper_gg_theme
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=\textwidth,keepaspectratio=true]{../figures/stanford-risk-1}

\caption{\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals.}\label{fig:stanford-risk}
\end{figure}
\end{CodeChunk}

@@

This report was generated on 2020-09-16 10:30:17 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.7.0               dotwhisker_0.5.0         
#>  [3] timereg_1.9.7             glue_1.4.2               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.2              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.1.2               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.9          
#> [15] visreg_2.7.0              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.2-3           
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.2.1          glmnet_4.0-2             
#> [25] Matrix_1.2-18             data.table_1.13.0        
#> [27] cowplot_1.1.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_2.0.1     base64enc_0.1-3    
#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
#> [61] evaluate_0.14       lattice_0.20-41     labeling_0.3       
#> [64] htmlwidgets_1.5.1   cmprsk_2.2-10       tidyselect_1.1.0   
#> [67] deSolve_1.28        plyr_1.8.6          R6_2.4.1           
#> [70] generics_0.0.2      Hmisc_4.4-1         multcomp_1.4-13    
#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
#> [76] haven_2.3.1         foreign_0.8-80      mgcv_1.8-33        
#> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
#> [85] grid_4.0.2          readxl_1.3.1        blob_1.2.1         
#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
#> [91] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
#> [94] viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Local:    master /Users/maxturgeon/Documents/git_repos/cbpaper
#> Remote:   master @ origin (git@github.com:sahirbhatnagar/cbpaper.git)
#> Head:     [73c0393] 2020-09-15: update CS3
\end{CodeOutput}
\end{CodeChunk}
